---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# multid

<!-- badges: start -->
[![Travis build status](https://app.travis-ci.com/vjilmari/multid.svg?branch=main)](https://app.travis-ci.com/vjilmari/multid/)
<!-- badges: end -->

The goal of multid is to provide tools for regularized measurement of multivariate differences between two groups (e.g., sex differences). Regularization via logistic regression variants enables inclusion of large number of correlated variables in the multivariate set while providing k-fold cross-validation and regularization to avoid overfitting. 

Predictive approach as implemented with regularized methods also allows for examination of group-membership probabilities and their distributions across individuals. In the context of statistical predictions of sex, these distributions are an updated variant to gender-typicality distributions used in gender diagnosticity methodology [(Lippa & Connelly, 1990)](https://doi.org/10.1037/0022-3514.59.5.1051).

Studies in which these methods have been used:

1. [Lönnqvist & Ilmarinen (2021). Using a Continuous Measure of Genderedness to Assess Sex Differences in the Attitudes of the Political Elite. Political Behavior.](https://doi.org/10.1007/s11109-021-09681-2)
2. [Ilmarinen et al. (2021). Is There a g-factor of Genderedness? Using a Continuous Measure of Genderedness to Assess Sex Differences in Personality, Values, Cognitive Ability, School Grades, and Educational Track. Manuscript in review.](https://doi.org/10.31234/osf.io/j59bs)



## Installation

You can install the released version of multid from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("multid")
```

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("vjilmari/multid")
```
## Example

This example shows how to measure standardized multivariate (both Sepal and Petal dimensions, four variables in total) distance between setosa and versicolor Species in iris dataset. 

```{r example1}
library(multid)
set.seed(91237)

D_regularized(
  data = iris[iris$Species == "setosa" | iris$Species == "versicolor", ],
  mv.vars = c(
    "Sepal.Length", "Sepal.Width",
    "Petal.Length", "Petal.Width"
  ),
  group.var = "Species",
  group.values = c("setosa", "versicolor")
)$D

# Use different partitions of data for regularization and estimation
D_regularized(
  data = iris[iris$Species == "setosa" |
    iris$Species == "versicolor", ],
  mv.vars = c(
    "Sepal.Length", "Sepal.Width",
    "Petal.Length", "Petal.Width"
  ),
  group.var = "Species",
  group.values = c("setosa", "versicolor"),
  size = 35,
  out = TRUE
)$D
```

This example first generates artificial multi-group data which are then used as separate data folds in the regularization procedure following separate predictions made for each fold. 

```{r example2}
# generate data for 10 groups
set.seed(34246)
n1 <- 100
n2 <- 10
d <-
  data.frame(
    sex = sample(c("male", "female"), n1 * n2, replace = TRUE),
    fold = sample(x = LETTERS[1:n2], size = n1 * n2, replace = TRUE),
    x1 = rnorm(n1 * n2),
    x2 = rnorm(n1 * n2),
    x3 = rnorm(n1 * n2)
  )
#'
# Fit and predict with same data
round(D_regularized(
  data = d,
  mv.vars = c("x1", "x2", "x3"),
  group.var = "sex",
  group.values = c("female", "male"),
  fold.var = "fold",
  fold = TRUE,
  rename.output = TRUE
)$D,2)
#'

# Different partitions for regularization and estimation for each data fold. Request probabilities of correct classification (pcc) and area under the receiver operating characteristics (auc) for output.

round(D_regularized(
  data = d,
  mv.vars = c("x1", "x2", "x3"),
  group.var = "sex",
  group.values = c("female", "male"),
  fold.var = "fold",
  size = 17,
  out = TRUE,
  fold = TRUE,
  rename.output = TRUE,
  pcc = TRUE,
  auc = TRUE
)$D,2)


```

