---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# multid

<!-- badges: start -->
[![Travis build status](https://app.travis-ci.com/vjilmari/multid.svg?branch=main)](https://app.travis-ci.com/vjilmari/multid/)
<!-- badges: end -->

The goal of multid is to provide tools for regularized measurement of multivariate differences between two groups (e.g., sex differences). Regularization via logistic regression variants enables inclusion of large number of correlated variables in the multivariate set while providing k-fold cross-validation and regularization to avoid overfitting. 

Studies in which the method has been used:

1. [Lönnqvist & Ilmarinen (2021). Using a Continuous Measure of Genderedness to Assess Sex Differences in the Attitudes of the Political Elite. Political Behavior.](https://doi.org/10.1007/s11109-021-09681-2)
2. [Ilmarinen et al. (2021). Is There a g-factor of Genderedness? Using a Continuous Measure of Genderedness to Assess Sex Differences in Personality, Values, Cognitive Ability, School Grades, and Educational Track. Manuscript in review.](https://doi.org/10.31234/osf.io/j59bs)



## Installation

You can install the released version of multid from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("multid")
```

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("vjilmari/multid")
```
## Example

This is a basic example which shows you how to measure standardized multivariate (both Sepal and Petal dimensions, four variables in total) distance between setosa and versicolor Species in iris dataset. Similar example is provided with artificially generated multi group data which are used as separate data folds in the procedure. 

```{r example}
library(multid)
set.seed(91237)

D_regularized(
  data = iris[iris$Species == "setosa" | iris$Species == "versicolor", ],
  mv.vars = c(
    "Sepal.Length", "Sepal.Width",
    "Petal.Length", "Petal.Width"
  ),
  group.var = "Species",
  group.values = c("setosa", "versicolor")
)$D

# Use different partitions of data for regularization and estimation
D_regularized(
  data = iris[iris$Species == "setosa" |
    iris$Species == "versicolor", ],
  mv.vars = c(
    "Sepal.Length", "Sepal.Width",
    "Petal.Length", "Petal.Width"
  ),
  group.var = "Species",
  group.values = c("setosa", "versicolor"),
  size = 35,
  out = TRUE
)$D

# multigroup model where the group variable is defined as fold.var, and output is produced separately for each group/fold. Groups are also used as folds in the k-fold cross-validation procedure.

# separate sample folds
# generate data for 10 groups
set.seed(34246)
n1 <- 100
n2 <- 10
d <-
  data.frame(
    sex = sample(c("male", "female"), n1 * n2, replace = TRUE),
    fold = sample(x = LETTERS[1:n2], size = n1 * n2, replace = TRUE),
    x1 = rnorm(n1 * n2),
    x2 = rnorm(n1 * n2),
    x3 = rnorm(n1 * n2)
  )
#'
# Fit and predict with same data
round(D_regularized(
  data = d,
  mv.vars = c("x1", "x2", "x3"),
  group.var = "sex",
  group.values = c("female", "male"),
  fold.var = "fold",
  fold = TRUE,
  rename.output = TRUE
)$D,2)
#'

# Different partitions for regularization and estimation for each data fold. Add probabilities of correct classification (pcc) as well.

round(D_regularized(
  data = d,
  mv.vars = c("x1", "x2", "x3"),
  group.var = "sex",
  group.values = c("female", "male"),
  fold.var = "fold",
  size = 17,
  out = TRUE,
  fold = TRUE,
  rename.output = TRUE,
  pcc = TRUE
)$D,2)


```

